\documentclass[11pt,a4paper]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{tikz}
\usepackage{url}

\pgfplotsset{
    unbounded coords=discard,
    filter discard warning=false,
    compat=1.18
}

% Activate high-performance PGFPlots caching framework
\usepgfplotslibrary{external}
\tikzexternalize[prefix=tikz-cache/]

% Custom conditional plotting filter definition macro
\pgfplotsset{
    discard if not/.style 2 args={
        filter point/.code={
            \edef\val{\thisrow{#1}}
            \edef\checkval{#2}
            \ifx\val\checkval
            \else
                \pgfmapsetdiscardpoint
            \fi
        }
    }
}

\title{\textbf{SpectralBL Multi-Campaign Analysis:}\\
       \large Cross-Ecosystem Stable Boundary Layer Comparison}
\author{\textbf{Automated Analytics Pipeline Engine}\\
        \small SpectralBL-Analytics Framework}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This report provides a comparative, low-rank structural analysis of stable boundary layer (SBL) regimes across multiple field campaigns and ecosystems. By projecting high-dimensional physical profile measurements from diverse sources (CASES-99, GABLS3, AMERIFLUX, SMEAR, NEON, ICOS, SHEBA) onto a unified low-rank continuous phase space, we reconstruct the underlying attractor manifold and analyze how different environments and forcings shape boundary layer organization. This framework provides an information-theoretic approach to diagnosing boundary layer complexity and mapping regime transitions across heterogeneous datasets.
\end{abstract}

\section{Introduction and Physical Context}

Traditional boundary layer parameterizations rely heavily on local gradient profiles and standard Monin-Obukhov Similarity Theory (MOST). However, in the very stable boundary layer (vSBL), the coupling between the surface layer and the upper air column is frequently severed by the development of a strong thermal inversion. The atmosphere enters a regime dominated by radiative cooling, low-level jet (LLJ) development, global intermittency, and submesoscale internal gravity waves.

In these conditions, local turbulence metrics such as $\mathrm{Ri}_g$ become unreliable diagnostics. To track structural reconfigurations across diverse ecosystems and measurement networks, the SpectralBL pipeline extracts continuous spatial pattern modes of the boundary layer state. The resulting low-rank coordinates $(\eta_1, \eta_2, \eta_3)$ remain dynamically active throughout stable and super-critical conditions, providing comparative insights into how different environments organize their stability structures.

\newpage

% Injection hooks for Mustache-rendered sub-sections
\input{generated/attractor.tex}

\newpage

\input{generated/regime.tex}

\newpage

\input{generated/conclusions_and_diagnostics.tex}

\newpage

\input{generated/bifurcation_spectrum.tex}

\newpage

\appendix

\section{Pipeline Metadata}

\begin{itemize}
    \item Source code: \url{https://github.com/davidengland/SpectralBL-Analytics}
    \item Build timestamp: \today{}
    \item Supported campaigns: CASES-99, GABLS3, AMERIFLUX (60+ sites), SMEAR, NEON, ICOS, SHEBA
    \item Mathematical basis: p-FEM spectral projection and scale-invariant rank truncation
    \item Templating: Mustache -- active
    \item Figure cache: TikZ external compilation -- \texttt{tikz-cache/}
\end{itemize}

\end{document}
